---
title: "Capstone"
author: "Justin P"
date: "2025-06-05"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Capstone Project

Project Description:This R Markdown file is the main source of data cleaning and data analytics for the Google Data Analytics Capstone Project. It attempts to answer the question: How do annual members and casual riders use Cyclistic bikes differently? This is part of Track A from the Capstone Project portion of the course using the bike-share dataset, Cyclistic, available by Motivate International Inc. 


## Install packages (if needed)

```{r install packages, message=FALSE, include=FALSE}
#install.packages("ggplot2")
#install.packages("tidyverse")
#install.packages("tidyr")
#install.packages("readr")
#install.packages("dplyr")

```

## Load environment and relevant packages

```{r packages}
library(ggplot2)
library(tidyverse)
library(readr)
library(tidyr)
library(dplyr)


```

## Load Datasets

```{r}
Q1_2019_df <- read_csv("C:/Users/Justin/Documents/Coursera/Google Data Analytics Professional Certificate/Course 8/Case_Study1/Data/Divvy_Trips_2019_Q1 - Divvy_Trips_2019_Q1.csv")


Q1_2020_df <- read_csv("C:/Users/Justin/Documents/Coursera/Google Data Analytics Professional Certificate/Course 8/Case_Study1/Data/Divvy_Trips_2020_Q1 - Divvy_Trips_2020_Q1.csv")

```

### Look at Q1-2019

```{r Q1_2019 Data Frame}
head(Q1_2019_df)
```

Column names:

```{r Q1_2019 Column Names}
colnames(Q1_2019_df)
```

### Look at Q1-2020

```{r Q1_2020 Data Frame}
head(Q1_2020_df)
```

Column names:

```{r Q1_2020 Column Names}
colnames(Q1_2020_df)
```

## Data Cleaning

### Make 2019 & 2020 column names consistent

Column mapping {2019:2020}: {trip_id: ride_id, start_time: started_at, end_time: ended_at, from_station_id: start_station_id, from_station_name: start_station_name, to_station_id: end_station_id, usertype: member_casual, ride_length: ride_length, day_of_week: day_of_week}

Columns not present in both: Q1_2019: {bikeid, tripduration, gender, birthyear} \| Q1_2020: {rideable_type, start_lat, start_lng, end_lat, end_lng}

```{r Rename columns for consistency}
library(dplyr)

Q1_2020_df_renamed <- Q1_2020_df %>%
  rename(
    trip_id           = ride_id,
    start_time        = started_at, 
    end_time          = ended_at, 
    from_station_id   = start_station_id,
    from_station_name = start_station_name,
    to_station_id     = end_station_id,
    to_station_name     = end_station_name,
    usertype          = member_casual,
  )

colnames(Q1_2020_df_renamed)

```

```{r}
head(Q1_2020_df_renamed)
```

## Check the structure of the dataframes

```{r structure of Q1_2019}
str(Q1_2019_df)
```

```{r structure of Q1_2020}
str(Q1_2020_df_renamed)
```

## Join the two data files

### In R

First, the two dataframe columns must be aligned

```{r Get all relevant columns}
all_cols <- c("trip_id", "start_time", "end_time", "from_station_id", "from_station_name", "to_station_id", "to_station_name", "usertype")
all_cols
```

Get all relevant columns for 2019 and 2020

```{r Get all relevant columns for 2019 and 2020}
df_2019 <- Q1_2019_df[all_cols]
df_2020 <- Q1_2020_df_renamed[all_cols]
```

Add a year column to each dataframe to keep track of users per year

```{r Add the years 2019 and 2020 as a column, respectivitly to each dataframe}
df_2019["Year"] = 2019
df_2020["Year"] = 2020

```

Last adjustments, making sure columns are of the same type

```{r}
df_2019$trip_id <- as.character(df_2019$trip_id) # double to character

```

Finally combine the 2019 and 2020 data frames

```{r Combine 2019 and 2020}
combined_df <- bind_rows(df_2019, df_2020)
head(combined_df)
```

### In SQL

### Export data frames and move to SQL to join dataset

```{r}
write.csv(Q1_2019_df, "2019_partial_clean.csv", row.names = FALSE)
write.csv(Q1_2020_df_renamed, "2020_partial_clean.csv", row.names = FALSE)

```

### Move to SQL and join datasets

This code was run in BigQuery. And will not work with the current markdown.

Using BigQuery as an easy SQL server. Steps: 1) Create a new project. 2) Create a new dataset, call it bicycle. 3) Upload "2019_partial_clean.csv" and "2020_partial_clean.csv" to the bicycle data set as tables (2019, 2020). 4) Run the code below in BigQuery.

```{r}

#SELECT
#  CAST(trip_id AS STRING) AS trip_id, # recast trip_id from int64 to string
#  start_time,
#  end_time, 
#  CAST(from_station_id AS STRING) AS from_station_id, # make sure from_station_id is a string
#  from_station_name, 
#  CAST(to_station_id AS STRING) AS to_station_id, # make sure to_station_id is a string
#  to_station_name, 
#  usertype, 
#  2019 AS Year, # Add a "Year" column for all users for 2019. 
#  TIMESTAMP_DIFF(end_time, start_time, MINUTE) AS trip_duration # find the trip duration for each user 
#FROM `project-1-461316.bicycle.2019`

#UNION ALL # combine all rows and columns from both tables 

#SELECT
#  CAST(ride_id AS STRING) AS trip_id, # recast trip_id from int64 to string
#  started_at AS start_time, 
#  ended_at AS end_time, 
#  CAST(start_station_id AS STRING) AS from_station_id,  # make sure start_station_id is a string
#  start_station_name AS from_station_name, 
#  CAST(end_station_id AS STRING) AS to_station_id,  #  # make sure end_station_id is a string
#  end_station_name AS to_station_name, 
#  member_casual AS usertype, 
#  2020 AS Year, # Add a "Year" column for all users for 2019. 
#  TIMESTAMP_DIFF(ended_at, started_at, MINUTE) AS trip_duration # find the trip duration for each user 

#FROM `project-1-461316.bicycle.2020`

```

## Question: How do annual members and casual riders use Cyclistic bikes differently?

### Data Cleaning - advanced

Find all users and make sure each user has the same title of {"Subscriber": "member", "customer": "causal"}

```{r}
combined_df <- combined_df %>%
  mutate(usertype = case_when(
    usertype == "member" ~ "Subscriber",
    usertype == "casual" ~ "Customer",
    TRUE ~ usertype  # Keep other values unchanged
  ))
```

## Data Analysis

Find the number of Subscribers and Customers for 2019 and 2020

```{r}
# Summarize counts by usertype and year
summary_df <- combined_df %>%
  group_by(Year, usertype) %>%
  summarise(count = n(), .groups = "drop")

# Create the bar plot
ggplot(summary_df, aes(x = factor(Year), y = count, fill = usertype)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "User Type Counts for 2019 and 2020",
       x = "Year",
       y = "Number of Users",
       fill = "User Type") +
  theme_minimal()
```

## Find the average trip duration for Customer and Subscriber

```{r Find the trip duration}
combined_df$start_time <- as.POSIXct(combined_df$start_time)
combined_df$end_time <- as.POSIXct(combined_df$end_time)

combined_df$trip_duration <- combined_df$end_time - combined_df$start_time
combined_df$trip_duration <- as.numeric(combined_df$trip_duration, units = "mins")

```

```{r Make a bar plot of the trip duration of subscribers and customers}
# Summarize the average trip duration by usertype and year
summary_df <- combined_df %>%
  group_by(Year, usertype) %>%
  summarise(
    average_duration = mean(trip_duration, na.rm = TRUE),
    sem = sd(trip_duration, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(summary_df, aes(x = factor(Year), y = average_duration, fill = usertype)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = average_duration - sem, ymax = average_duration + sem),
    position = position_dodge(width = 0.9),
    width = 0.2
  ) +
  labs(
    title = "Average Trip Duration by User Type (2019 and 2020)",
    x = "Year",
    y = "Average Trip Duration (minutes)",
    fill = "User Type"
  ) +
  theme_minimal()
```

```{r}

# Calculate average and SEM
summary_df <- combined_df %>%
  group_by(Year, usertype) %>%
  summarise(
    average_duration = mean(trip_duration, na.rm = TRUE),
    sem = sd(trip_duration, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Plot with facet for usertype
ggplot(summary_df, aes(x = factor(Year), y = average_duration, fill = factor(Year))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = average_duration - sem, ymax = average_duration + sem),
    position = position_dodge(width = 0.9),
    width = 0.2
  ) +
  facet_wrap(~ usertype) +
  labs(
    title = "Average Trip Duration by Year for Each User Type",
    x = "Year",
    y = "Average Trip Duration (minutes)",
    fill = "Year"
  ) +
  theme_minimal()

```
